\label{sec:multi-port-inputs}%


Consider a special kind of an alphabet
$\Sigma = \langle \Sigma_1, \Sigma_2, \dots, \Sigma_k \rangle$ such
that $\Sigma_i \cap \Sigma_j = \emptyset$ for all $i, j \in
[1..k]$. We will call $\Sigma_i$ as the alphabet of \emph{port} $i$,
and $\Sigma$ as a \emph{multi-port} alphabet. We
look at DFAs over such multi-port alphabets. Such DFAs model systems
that listen to inputs from different processes and perform actions
based on them. Sometimes the order in which the system receives its
inputs from different ports is not relevant.%

For example, at a state
$s$, if the system receives $a$ from port 1 and $b$ from port 2, in any
order, then it has to go to state $t$.  A DFA would model this with
transitions: $s \xra{a} s_a \xra{b} t$ and $s \xra{b} s_b \xra{a}
t$. A DSA would contain two transitions $s \xra{ab, ba} t$ (and some
other transitions, if needed, to take care of $aa$, $bb$). To get a
more succinct notation we will use a $\parallel$ operator. We will
write $s \xra{a \parallel b} t$ to mean that at state $s$, when
both $a$ and $b$ are received, the automaton moves to $t$. When there
are several components, this notation leads to significant
succinctness, for instance $a_1 \parallel a_2 \cdots \parallel a_n$
stands for all the $n!$ permutations of $a_1$ to $a_n$. We will also
allow expressions of the form $a_1 a_2 \parallel b$, which stands for
the set of words $\{a_1 a_2 b, a_1 b a_2, b a_1 a_2\}$ which shuffles
$a_1a_2$ and $b$. We will now formalize this idea by enhancing DSA with
the $\parallel$ operator and then consider the problem of synthesizing
such extended DSA. We begin with some notation.%

% DFAs model
% a system consisting of $k$-components, where the $i^{th}$ component
% has sole control over actions of $\Sigma_i$. However every component
% can observe the actions generated by all other components. So, instead
% of modeling such a system as a network of automata, we consider it to
% be a single DFA over this multi-port alphabet. Therefore, for a start, 
% there is no ``independence'' between the ports: for example if
% action $a$ is controlled by process $1$ and action $b$ is controlled
% by process $2$, we cannot in general commute the sequence $ab$ with
% $ba$: it may be the case that process $2$ waits for process $1$ to
% emit an $a$. However, in the DFA, there could be
% specific parts where there is independence, in other words, there are
% states $s$, $t$ and a set of actions coming from different ports such
% that all interleavings among different ports are possible between $s$
% and $t$. We wish to extend the definition of a suffix-reading
% automaton with the aim to capture more succinctly these ``diamonds'' occurring in the
% DFA. %



\paragraph*{Notation} For a word $w \in \Sigma^*$, we write
$\proj{i}(w)$ for the projection of $w$ onto the set $\Sigma_i$. For
instance, if $\Sigma_1 = \{a_1, a_2\}, \Sigma_2 = \{b_1, b_2\}$ and
$w = a_1 b_1 a_2 a_1 b_2 b_2$, we have $\proj{1}(w) = a_1 a_2 a_1$ and
$\proj{2}{w} = b_1 b_2 b_2$.  We write $\partial w$ for the $k$-tuple
$(\proj{1}(w), \proj{2}(w), \dots, \proj{k}(w))$ of projections of $w$
onto each port $\Sigma_i$, and will call $\partial w$ the
\emph{decomposition} of $w$. Notice that if
$\partial w_1 = \partial w_2$ for two words $w_1, w_2$, then $w_1$ and
$w_2$ have the same order of events within each port, but could have a
different ordering between letters from different ports.%


\begin{definition} Let
 $\Sigma = \langle \Sigma_1, \Sigma_2, \dots, \Sigma_k \rangle$ be a
 multi-port alphabet. A \emph{multi-port (deterministic)
   suffix-reading automaton} (written \mdsa in short) $\Aa$ is a
 tuple $(Q, \Sigma, q^{init}, \delta, F)$ where $Q$, $q^{init}$ and
 $F$ are a finite set of states, the initial state and a set of
 accept states, respectively. The transition relation
 $\d \incl Q \times (\Sigma_1^* \times \Sigma_2^* \times \cdots
 \times \Sigma_k^*) \times Q$: each transition is of the form
 $(q, (u_1 \parallel u_2 \parallel \cdots \parallel u_k), q')$ where
 $u_i \in \Sigma_i^*$ (not all of them can be $\epsilon$).
 % We assume that for every two outgoing
 % transitions
 % $(q, (u_1, u_2, \dots, u_k), q_1)$ and
 % $(q, (v_1, v_2, \dots, v_k), q_2)$ there is at least one port $i$
 % such that $u_i$ and $v_i$ are suffix-incomparable, that is,
 % $u_i \not \sfx v_i$ and $v_i \not \sfx u_i$.
\end{definition}%

%\textcolor{red}{Jan 10, 2023. The first instinct is to say that the
% transition $q \xra{u_1 \parallel u_2 \cdots u_k} q'$ can be replaced
% with multiple transitions to get the usual DSA of the previous
% section. For instance $q \xra{a \parallel b} q'$ can be replaced
% with $q \xra{ab} q'$ and $q \xra{ba} q'$. But this is not correct in
% general!
% Suppose there is a third port. Consider a transition
% $q \xra{a \parallel b \parallel \epsilon} q'$. The third $\epsilon$
% is to say that we do not care what happens in that port. It does not mean
% that there is no input! We cannot just replace it with $\xra{ab}$
% and $\xra{ba}$. If we do so, the word $a c b$ at state $q$ would not
% move
% to $q'$ ($c$ is in third port), whereas it should. In fact, we
% cannot replace it with finitely many transitions: $a w b$ with $w$ a
% word using the third port alphabet is good for $a \parallel b
% \parallel \epsilon$. So, we need a completely fresh analysis for extending
% the parallel operation to DSA. We could keep this for future work.}%

\paragraph*{Semantics} $B=\langle B_1,B_2,\dots,B_{k'} \rangle$ is a tuple of tapes, one for each port and some additional ones ($k'\ge k$). Each tape has two tape heads, $L$ and $R$, that track the previous `match' and latest input respectively. 

\begin{enumerate}
\item All tapes are initialized at position 0, with the $L$ and $R$ heads both pointing there. (call the head positions for the $B_i$ tape, $L_i$ and $R_i$)

\item When an input $a \in \Sigma_i$ is seen, the tape $B_i$ has its $R$-head move right by one position and store the value $a$. ($R_i := R_i+1$)

\item A transition is \emph{matched} if every $B_i$ has $u_i$ as a suffix of its stored word, with either $L_i=R_i$ or the entire $u_i$ present between $L_i$ and $R_i$. Additionally, at least one $B_i$ has $R_i-L_i>0$.

\item A matched transition is then \emph{triggered}, that is $\forall u_i \ne \epsilon$, the $L$-head of $B_i$ moves to the position of its $R$-head. ($L_i := R_i$)
\end{enumerate}

The states of a multi-port DSA are defined as per the requirements of the system it models. For a system with internal variables $S_1,S_2,\dots$ with each variable $S_i$ having an alphabet $\Sigma_{S_i}$, the corresponding \mdsa has states $Q=\Sigma_1 \times \Sigma_2 \times \dots$. The additional tapes $B_{k+1}, B_{k+2}, \dots, B_{k'}$ are for the internal variables used in the system. These variables can read input as well as produce output, over the common alphabet representing their state space. %For example, a variable $S_i$ has an alphabet $\Sigma_{S_i}$ with letters that can be input as well as output.

Transitions must then be enhanced in our semantics i.e.  a transition $(q, (u_1, \dots, u_k), q')$ can now require the local variables $S_1, S_2,\dots$ to have values $(s_1, s_2 \dots)$  for it to match. When triggered, the values then change to $(s'_1,s'_2,\dots)$. Assume each variable $S_i$ has its corresponding tape labeled $B_{S_i}$ (we relabel the tapes $B_{k+1}, B_{k+2}, \dots, B_{k'}$ appropriately). Let us use the syntax $\langle (q, (u_1, \dots, u_k), q') , (s_1, s_2 \dots), (s'_1,s'_2,\dots) \rangle$ to represent this.

\begin{enumerate}
\item A transition $\langle (q, (u_1, \dots, u_k), q') , (s_1, s_2 \dots), (s'_1,s'_2,\dots) \rangle$ is matched if each $B_i (i\le k)$ has $u_i$ as suffix, the last letter of each $B_{S_i}$ is $s_i$, and either $L_i=R_i$ or the entire $u_i$ present between $L_i$ and $R_i$. Additionally, at least one tape is non-empty between its $L$ and $R$ heads.

\item A matched transition is then triggered, moving  the $L$-head of $B_i$ (or $B_{S_i}$) to the position of its $R$-head $\forall u_i \ne \epsilon$ and $\forall s_i \ne \epsilon$. Additionally $\forall s'_i \ne \epsilon$, the corresponding $B_{S_i}$ moves its $R$-head by one position to the right and stores $s'_i$.
\end{enumerate}

% Given states $q, q'$, we say that a word $w \in \Sigma^*$ matches a
% transition $(q, (u_1, \dots, u_k), q')$ at $q$ if
% $u_i \sfx \proj{i}(w)$ for all $i$. When the automaton is in state
% $q$ and reads a word $w$, we say that transition
% $\theta = (q, (u_1, \dots, u_k), q')$ is \emph{triggered} if $w$
% matches $\theta$ at $q$, and moreover, no proper prefix of $w$
% matches any outgoing transition at $q$.

 For a word $w \in \Sigma^*$, the run of $\Aa$ on $w$ is given by:
 \begin{align*}
   q^{init} = q_0 \xra{w_1} q_1 \xra{w_2} \cdots \xra{w_{k}} q_k \xra{w_{k+1}}
 \end{align*}
 such that $w = w_1 w_2 \dots w_kw_{k+1}$ and there are transitions
 between $q_{i-1}$ and $q_i$ triggered on $w_i$ at $q_{i-1}$, for
 every $1 \le i \le k$. The run is \emph{accepting} if
 $w_{k+1} = \epsilon$ (no dangling letters) and $q_k$ is accepting. A
 word $w$ is accepted by SRA $\Aa$ if it has an accepting run. As
 usual, we define the language $\Ll(\Aa)$ to be the set of words that
 are accepted. Notice that we have extended the semantics of
 deterministic suffix-reading automata to the multi-port
 case. Henceforth we will write mDSA for multi-port DSA.%

\textcolor{red}{TODO: Semantics for Row sequences}

\begin{lemma}
 A multi-port DSA has a unique run over every word.
\end{lemma}%

\begin{theorem}
 For every DFA $M$ over a multi-port alphabet $\Sigma$, there exists
 a multi-port DSA $S_M$ such that $\Ll(M) = \Ll(S_M)$, and
 vice-versa.
\end{theorem}
\begin{IEEEproof} (Sketch.) For DFA to SRA: first complete the DFA, and
 then replace $q \xra{a} q'$ with
 $q \xra{(\e, \e, \dots, a, \dots, \e)} q'$ putting $a$ at the right
 port.%

 For SRA to DFA: replace $(u_1, u_2, \dots, u_k)$ with several
 transitions $q \xra{w} q'$ one for each $w$ such that
 $\partial w = (u_1, \dots, u_k)$. This gives an SRA. From previous
 theorem that SRA can be converted to DFA, we get DFA.%

\end{IEEEproof}%



\bibliographystyle{plain}
\bibliography{main-edt-synthesis}
\end{document}
