\section{Introduction}
\label{sec:intro}

  % Automata are very useful. In particular in formal verification and model-based testing.

  Finite state automata are all-pervasive, adopting different roles in different contexts. In this work, we focus on the role of automata as \emph{formal models of system requirements} and their subsequent application in Model Based Testing (MBT). MBT is a paradigm in formal methods where the requirements of a system under test (SUT) are first converted to a formal model. This formal model is then used to automatically generate test cases for the SUT. There are many MBT tools (see \cite{10.1145/1353673.1353681,10.1002/stvr.456} for a survey) and in several of them, the formal model is given by some extension of finite automata. 

  %Automata can be made concise.Making automata concise is a long line of work. Different ways in which they could be made concise. In MBT this helps in writing more concise specifications. 

  Converting requirements to a Deterministic Finite Automaton (DFA) or a Non-deterministic Finite Automaton (NFA) is often cumbersome. This is because DFAs and NFAs describe low level computations which reason about individual actions. There are various ways in which automata can be made consise. In Generalized Automata~\cite{DBLP:books/lib/Eilenberg76,GIAMMARRESI1999191}, transitions contain words instead of single actions. A transition $q \xra{abb} q'$ is fired from $q$ if the word $abb$ is after the automaton reaches $q$. Therefore requirements which depend on an exact sequence of actions can be represented using generalized automata more concisely. A natural extension is to allow for regular expressions in transitions, resulting in the model of Expression Automata~\cite{10.1007/978-3-540-30500-2_15} and its deterministic variant. Deterministic Expression Automata (DEA) impose restrictions on the regular expressions that can be used: they should result in a prefix-free language, and moreover, every pair of regular expressions out of state is language disjoint. This results in DEA recognizing a subset of regular languages. On the other hand, in Expression Automata, the transitions are so powerful that the states have almost no meaning: the entire automaton can be reduced to a single state and a single transition labeled by the regular expression representing the language. For languages leading to big automata, the corresponding regular expression can be complicated and incomprehensible. The goal is to hit a sweet spot where the transition labels are not too complicated, and at the same time there are fewer states than a DFA. 

   

  Deterministic suffix-reading automata (DSAs)~\cite{DBLP:journals/corr/abs-2410-22761} are a recently introduced automaton model for regular languages. Transitions are  labeled with words (as in a generalized automaton), however, the semantics is different. A transition $q \xra{abb} q'$ is fired from $q$ if a word \emph{ending} with the suffix $abb$ is seen. When there are multiple transitions out of a state, the first transition that matches is fired. Figure~\ref{fig:dsa-examples} gives some examples of DSAs.

  \begin{figure}
  \centering
  \begin{tikzpicture}[state/.style={circle, inner sep=2pt, draw, thick}]
    \begin{scope}[every node/.style={state}]
      \node [double] (0) at (0,0) {\scriptsize $q_0$};
    \end{scope}
    \begin{scope}[->, >=stealth, thick]
      \draw (-0.5, 0) to (0);
      \draw (0) to [loop above] node {\scriptsize $ab$} (0);
    \end{scope}
    \node at (0, -0.7) {\scriptsize $(a)$};

    \begin{scope}[xshift=2cm]
      \begin{scope}[every node/.style={state}]
        \node (0) at (0,0) {\scriptsize $q_0$};
        \node [double] (1) at (1.5, 0) {\scriptsize $q_1$};
      \end{scope}
      \begin{scope}[->, >=stealth, thick]
        \draw (-0.5, 0) to (0);
        \draw (0) to [loop above] node  {\scriptsize $ba$} (0);
        \draw (0) to node [above] {\scriptsize $ab$} (1);
      \end{scope}
      \node at (0.54, -0.7) {\scriptsize $(b)$};
    \end{scope}

      \begin{scope}[xshift=6cm]
        \begin{scope}
          \node [rectangle, rounded corners, draw, inner sep=3pt, thick] (0) at (0,0) {\scriptsize \texttt{Alarm=Off}};
          \node [rectangle, rounded corners, draw, inner sep=3pt, thick] (1) at (0,1.5) {\scriptsize \texttt{Alarm=On}};
          \node [rectangle, rounded corners, draw, inner sep=3pt, thick, double] (2) at (4, 0) {\scriptsize \texttt{Error}};
        \end{scope}

        \begin{scope}[->, >=stealth, thick]
          \draw (-1,0) to (0);
          \draw (0) to [bend left] node [left] {\scriptsize \texttt{On}} (1);
          \draw (1) to [bend left] node [right] {\scriptsize \texttt{Off}} (0);
          \draw (0) to node [above] {\scriptsize \texttt{PanicSw;PanicSw}} (2);
        \end{scope}
        \node at (2, -0.7) {\scriptsize $(c)$};
      \end{scope}
    
  \end{tikzpicture}
  \caption{Examples of Deterministic Suffix-reading Automata (DSA)}
  \label{fig:dsa-examples}
  \end{figure}

  In Figure~\ref{fig:dsa-examples}(a), the DSA accepts all words that end with $ab$, for instance $bab, bbab, babab$, etc. The transition $q_0 \xra{ab} q_0$ is fired as soon as a word ending with $ab$ is seen. In Figure~\ref{fig:dsa-examples}(a), we can view the state $q_0$ as tracking two suffixes $ba$ and $ab$. When $ba$ is seen before $ab$, the self-loop is triggered; otherwise when $ab$ is seen first, the transition to $q_1$ is taken. For example, on word $bbab$, the automaton starts at $q_0$, and on $bba$ it fires $q_0 \xra{ba} q_0$, reads $b$ and remains there. Since $q_0$ is non-accepting, the word is rejected. However, on $bbaab$, the run loops on $bba$ back to $q_0$, and then reads the next $ab$ to reach $q_1$, and accept. The final example Figure~\ref{fig:dsa-examples}(c) is a \emph{suffix-based specification} of a part of an automotive software: when the alarm is off, and the panic switch is pressed twice within one second, flag an error. Here, we assume there is a symbol \texttt{tick} to denote the elapse of $1$ unit of time. So, on a word \texttt{tick}; \texttt{On}; \texttt{tick} \texttt{Off}; \texttt{PanicSw}; \texttt{PanicSw}, an error should be flagged (we have added a separator symbol ; just for clarity). This specification is succinctly represented by the DSA in Figure~\ref{fig:dsa-examples}(c).

   Since transitions in a DSA wait for specific patterns to appear, DSAs are able to represent suffix-based specifications succinctly. When we consider systems with multiple components (which we call ports), the requirements can talk about different combinations of inputs. Here are two examples. 

   \begin{figure}
   \centering
   \begin{tikzpicture}
      \begin{scope}[every node/.style={rectangle, rounded corners, draw, inner sep=3pt, thick}]
        \node (0) at (0,0) {\scriptsize \texttt{Off}};
        \node [double] (1) at (0,1.5) {\scriptsize \texttt{Error}};
        \node (2) at (3.2,0) {\scriptsize \texttt{Started}};
        \node (3) at (3.2,1.5) {\scriptsize \texttt{Drive}};
      
      \end{scope}

      \begin{scope}[->,>=stealth, thick, auto]
        \draw (-0.75,0) to (0);
        %\draw (0) to [bend left] node [left] {\scriptsize \texttt{On}} (1);
        %\draw (1) to [bend left] node [right] {\scriptsize \texttt{Off}} (0);
        \draw (0) to node [below] {\scriptsize $B \parallel P \parallel$ \texttt{On}} (2);
        \draw (2) to node [right] {\scriptsize $B \parallel D$ } (3);
        \draw (3) to node [left] {\scriptsize $B \parallel P \parallel$ \texttt{Off}$~$} (0); 
        \draw (3) to node [above] {\scriptsize $D \parallel$ \texttt{Off}} (1);
      \end{scope}

      \node at (1.5, -1.2) {\scriptsize $(a)$};

      \begin{scope}[xshift=6.5cm]
        \begin{scope}[every node/.style={rectangle, rounded corners, draw, inner sep=3pt, thick}]
          \node [double] (0) at (0,0.75) {\scriptsize \texttt{PhoneOff}};
          \node [double] (1) at (3,0.76) {\scriptsize \texttt{PhoneOn}};
        \end{scope}

        \begin{scope}[->,>=stealth, thick, auto]
          \draw (-1,0.75) to (0);
          \draw (0) to [bend left] node {\scriptsize $F \parallel T_1 T_2 T_3 \parallel$ \texttt{Press}} (1);
          \draw (1) to [bend left] node {\scriptsize \texttt{Press}} (0);
        \end{scope}

        \node at (1.5, -1.2) {\scriptsize $(b)$};


      \end{scope}
   \end{tikzpicture}
   \caption{Concurrent suffix-based specifications as multi-port DSAs}
   \label{fig:mdsa-examples}
   \end{figure}
   \paragraph*{Car Security System.} For a car to start, you need: key in ignition ($K$), brake pedal pressed ($B$), transmission in park ($P$), and no alarm triggered. The actions $K$, $B$ and $P$ can occur in any order. A DSA would need to represent this requirement using all $6$ permutations. Instead, we assume an alphabet that is distributed across different \emph{ports} -- ignition, brake pedal, transmission, and denote the concurrent actions as $K \parallel B \parallel P$. Figure~\ref{fig:mdsa-examples}(a) illustrates the enriched DSA.

   \begin{itemize}
   \item When the brake pedal is pressed ($B$) and transmission is in park $(P)$, if ignition becomes on ($ON$), then the car is started.
   \item If the brake pedal is pressed ($(B)$) and the transmission is moved to drive ($D$), the car goes to the drive mode.
   \item When the brake pedal is pressed ($B$), transmission is in park, and the ignition becomes off ($OFF$), the car is switched off. If ignition is switched off when brake is not pressed or when transmission is in drive/reverse, signal an error.
   \end{itemize}

  
   \paragraph*{Smartphone Lock Pattern} To unlock a phone with pattern: fingerprint verification ($F$) and drawing specific gesture with three touch points ($T_1;T_2;T_3$). This is represented as $F \parallel (T_1 T_2 T_3)$ and the enriched DSA is shown in Figure~\ref{fig:mdsa-examples}(b).

   \begin{itemize}
   \item When the button is pressed ($B$), fingerprint verified and a correct pattern is received $(F_1 F_2 F_3)$, then the phone is switched on.
   \item If the phone is on, and the button is pressed ($B$), the phone is switched off.  
   \end{itemize}
   
  In Appendix~\ref{app:examples-of-parallel}, we give many more examples of similar situations that involve multiple ports and suffix-based requirements across all of them. To the best of our knowledge, no automaton model can represent such specifications succinctly. Incorporating concurrency into the specification is paramount when requirements talk about different components of the distributed system. 
  
  \subsubsection*{Contributions.} The goal of this paper is to investigate the
  addition of a $\parallel$ operator coupled with output to ports into the
  transition syntax of DSAs. We call the resulting model as \emph{multi-port
  DSAs}, where the alphabet is distributed (partitioned) across multiple ports.
  The challenge lies in describing an operational semantics that matches with
  our intentions. As our first technical contribution, we provide a formal
  operational semantics for multi-port DSAs. 
  
  We then show that state reachability (or emptiness) for mDSAs is PSPACE-complete. For the simpler models DFAs or DSAs, reachability is linear in the input size. The higher complexity for mDSAs ties with the intuition that mDSAs can be exponentially more succinct. Our proof of PSPACE-hardness indeed exhibits such an example. 

  As our next contribution, we apply the mDSA framework to provide a formal semantics for Expressive Decision Tables (EDTs), a notation introduced by \cite{DBLP:conf/date/VenkateshSKA14} to incorporate both state-based and stream-based requirements in one convenient tabular form. The EDT notation has been successfully deployed in many industrial projects. In \cite{DBLP:conf/date/VenkateshSKA14} it has also been observed that the time and effort involved in writing EDT specifications was considerably low compared to other formalisms like  Statecharts. Many test generation tools for EDT specifications have been developed: given an EDT, and a row, the tool can generate a sequence of input actions that can trigger the input-side conditions of the particular row~\cite{DBLP:conf/enase/VenkateshSZA15a,DBLP:conf/icst/AgrawalVSZV20}. %The tricky part in the EDT notation is the use of input-output (I/O) variables that make multiple rows dependent. 
  Although EDTs have been used extensively for test generation, the notation does not yet have a formal operational semantics because of a complex interplay between concurrency and suffix-based rules. We close these gaps using the mDSA technology: we consider a fragment of the EDT notation, and map each EDT to an mDSA. This in turn gives an operational semantics for the EDT. Moreover, test generation for EDTs reduces to state reachability in mDSAs. 
  
  
  
  

  

  % Contributions. 
  % - Add a parallel operator to the syntax. Examples 2 and 3. First time that both suffix-based and concurrency have been incorporated.
  % - Challenge lies in describing an operational semantics that matches with our intentions.
  % - State reachability of mDSAs (emptiness) is PSPACE-complete. Compare with DFAs/DSAs where it is polynomial-time. This ties with the intuition that mDSAs can be exponentially more succinct.
  % - Application: EDT to mDSAs. Test generation for EDTs as emptiness of mDSAs.
  
  % Why easy-to-write specifications are important? Why formal semantics is important? 
  
  % EDT introduction. Some examples of EDTs. 

  % An intricate EDT example for which no current tool can generate a test case? 

  % Contributions

  % 2 to 3 pages
