\documentclass[a4paper, 11pt]{llncs}

%\documentclass[conference]{IEEEtran}
%\IEEEoverridecommandlockouts

\usepackage{amsmath,amssymb}
\usepackage{stmaryrd}
\usepackage{tikz}
\usepackage{url}
\usetikzlibrary{automata, positioning, arrows}
\usepackage{array}
%\ifCLASSOPTIONcompsoc
% requires cite.sty v4.0 or later (November 2003)
%\usepackage[nocompress]{cite}
%\else
%\usepackage{cite}
%\fi

%% Macros

% Greek
%\newtheorem{definition}{Definition}


%\newtheorem{example}{Example}
%\newtheorem{lemma}{Lemma}
%\newtheorem{corollary}{Corollary}
%\newtheorem{proof}{IEEEproof}
%\newtheorem{theorem}{Theorem}

\renewcommand{\a}{\alpha}
\renewcommand{\b}{\beta}
\newcommand{\s}{\sigma}
\renewcommand{\d}{\delta}
\newcommand{\e}{\epsilon}
\newcommand{\m}{\mu}

\newcommand{\D}{\Delta}

\newcommand{\rr}{\mathsf{r}}
\newcommand{\bb}{\mathsf{b}}
\renewcommand{\gg}{\mathsf{g}}
\newcommand{\yy}{\mathsf{y}}
\newcommand{\ww}{\mathsf{w}}
% Mathcal
\newcommand{\Aa}{\mathcal{A}}
\newcommand{\Ll}{\mathcal{L}}
\newcommand{\Oo}{\mathcal{O}}
\newcommand{\Ss}{\mathcal{S}}
\newcommand{\Ww}{\mathcal{W}}
% Symbols
\newcommand{\imp}{\Rightarrow}
\newcommand{\xra}{\xrightarrow}
\newcommand{\Land}{\bigwedge}
\newcommand{\incl}{\subseteq}

% Misc
\newcommand{\ssczo}{C_{0 \to 1}}
\newcommand{\coz}{C_{1 \to 0}}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
%\renewcommand{\sp}{\operatorname{SP}}
%\renewcommand{\sp}{\sigma}
\renewcommand{\sp}{\mathcal{S}\mathcal{L}}
\newcommand{\context}{\mathcal{C}}
\newcommand{\border}{\mathcal{B}}
\newcommand{\nbd}{(\mathcal{C} + \mathcal{B})}
\newcommand{\prfx}{\sqsubseteq_{\mathsf {pr}}}
\newcommand{\pprfx}{\sqsubset_{\mathsf {pr}}}
\newcommand{\sfx}{\sqsubseteq_{\mathsf {sf}}}
\newcommand{\psfx}{\sqsubset_{\mathsf {sf}}}
\newcommand{\proj}[1]{\mathsf{P}_{#1}}
\newcommand{\spm}{\sp^{\mathsf{m}}}
\newcommand{\sfxm}{\sfx^{\mathsf{m}}}
\newcommand{\psfxm}{\psfx^{\mathsf{m}}}
%\renewcommand{\IEEEQED}{\IEEEQEDopen}

\newcommand{\PSPACE}{\operatorname{\textsc{Pspace}}}
\newcommand{\NP}{\mathsf{NP}}
\newcommand{\coNP}{\mathsf{coNP}}
\newcommand{\vertex}{\textsc{VertexCover}~}
\newcommand{\mdsa}{DSA$^{\mathsf{m}}$}

\newcommand{\out}{\operatorname{Out}}
\newcommand{\outp}{\operatorname{Out}_{\sqsubset}}

\newcommand{\lsfx}[1]{\sqsubseteq_{\scriptscriptstyle
    \mathsf{lsf}}^{\scriptscriptstyle #1}}
\newcommand{\sub}{\sqsubseteq_{\mathsf{sub}}}
\newcommand{\rsub}{\sqsubset_{\mathsf{sub}^{\mathsf{r}}}}
\renewcommand{\e}{\varepsilon}
\renewcommand{\epsilon}{\e}

\newcommand{\Tt}{\mathcal{T}}
\newcommand{\Ii}{\mathcal{I}}
\newcommand{\Vv}{\mathcal{V}}
%\input{macros}
\newcommand{\specialcell}[2][c]{%
  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
\newcommand{\var}{\operatorname{Var}}
\newcommand{\dom}{\ensuremath{dom}}


\begin{document}
\title{Suffix-reading automata in a practical setting}

\maketitle

\input{definition.tex}

\input{edt.tex}

\section{Test generation}
\label{sec:test-generation}

% What is test generation? The decision problem
Given an EDT $\Tt$ and a row $R$ of $\Tt$ as input, the \emph{test
  generation problem} is to decide if there exists an execution
$\pi \in \Ll(\Tt)$ such that $R$ is triggered in $\pi$.
%
The problem is
intrinsically difficult: one might have to match an intermediate row
multiple times in order to generate a test case for $R$, in fact this
number could even be exponential. This is witnessed in the EDT that
implements an $n$ bit counter. In order to match a row with $n$ ones,
the row with $0$ followed by $(n-1)$ $1$ needs be matched $2^{n-1}$
times. 
%
But thanks to the semantics of EDTs as Mealy machines, a natural
algorithm for this problem is to compute the Mealy machine 
on the fly and detect whether the row is triggered by a simple
traversal of the machine.


\begin{theorem}
  \label{thm:test-generation-PSPACE}
  Test generation for EDTs is $\PSPACE$-complete.
\end{theorem}

We prove below that test generation is $\PSPACE$-complete

For an EDT, its mDSA has exponentially many
configurations of tapes. But each configuration requires polynomially many bits (the tapes themselves). Detecting a test
case amounts to non-deterministically guessing a sequence of configurations. By standard complexity arguments, this gives a
$\PSPACE$-upper bound. %As we saw, all other EDTs can be converted in polynomial time to basic EDTs.

For the hardness, we reduce the problem of DFA
intersection. Consider $n$ DFA $\Aa_1, \Aa_2, \dots, \Aa_n$. The
problem is to decide if their intersection is empty. Each $\Aa_i$
can be considered as an EDT with row sequences. Assume that the sets
of states are mutually disjoint. Add a binary local variable $l_i$ and rules
such that $l_i = 1$ if ``current'' state of $\Aa_i$ is accepting and
otherwise $l_i = 0$. Add a row $\Land_i l_i = 1 \imp U$ for some
update. This row matches iff there is a non-empty intersection.

\section{Todos.}

\begin{itemize}
\item Compare the succinctness of mDSAs with DSAs. Understanding the
  effect of partitioning the alphabet and the automaton having the
  ability to treat them as independent events. We expect exponential
  succinctness easily, in terms of the number of input variables.

\item Comparison of mDSAs with streaming string transducers. Seem
  incomparable - the motivations are different. What about transducers
  with regular look ahead? 

\item Reachability of mDSAs: algorithm and a potential encoding in
  SAT?
  
\item EDTs to mDSAs: formal semantics for EDTs as mDSA. 

\item Test generation algorithm for EDTs reduced reachability in
  mDSAs. Complexity of test generation. 

\item Can we encode the test generation procedure as a SAT solving
  algorithm, instead of explicitly constructing the EDT and then
  running a reachability algorithm? This SAT-based solution should
  take an EDT and spit out a bunch of SAT queries, to cover each
  row. This also allows us to ask for ``bounded'' test case
  generation. 

\item Experiments. Give arguments to show that situation is better
  than before. Independently, is mDSA useful in a different
  setting. Potentially a new case study that illustrates the use of
  EDT methodology in a new setting.  
  
\end{itemize}

\bibliographystyle{splncs04}
\bibliography{mDSA}

\end{document}	
