\subsubsection*{Related work.}
Use of automata to model distributed systems has been explored widely.
Esterel~\cite{DBLP:journals/scp/BerryG92} is a synchronous programming language
that provides a parallel composition operator. Harel's
Statecharts~\cite{DBLP:journals/scp/Harel87} also support parallel composition
of state transition systems. In both these languages the reaction of a system
in response to inputs on each port can be specified separately and the
different specifications can be composed using the parallel operator. A system
model can then refer to the states of each port's specification to model the
system behaviour. In mDSAs, we do not compose automata. Instead there is a single automaton, but the transitions include a parallel composition operator. More importantly, mDSA allows for suffix-based event triggers. %specifications are more succinct than Statecharts or Esterel specifications of System behaviours similar to the examples described in this
%paper. 
Petri nets~\cite{CAPetri}
have been studied extensively to model concurrent systems. In Petri nets each
letter of each port's alphabet will have to be represented using a place and
the places combined to model the system. There is no mechanism to include suffix-based triggers. %This will not be as succinct as an
%mDSA. 
The work that is closest to ours is Input/Output Partial Order
Automata (IOPOA)~\cite{10.5555/2391293.2391305}. In their work transitions
execute non-atomically reacting to asynchronous inputs on several ports. The
main differences between IOPOA and mDSA: in mDSA transitions react
atomically, transition labels can be strings representing a suffix for each port and not just letters,
and different transitions may trigger on the same string at a port. For example consider the mDSA in Figure~\ref{fig:mdsa-examples}(a), the transition from $\mathtt{Off}$ to $\mathtt{Started}$ and $\mathtt{Started}$ to $\mathtt{Drive}$ can be taken on the same input $B$. In an IOPOA this will require two occurences of $B$. IOPOA was primarily designed to enable test generation whereas the aim of mDSAs is succinct specification of requirements.

\subsubsection*{Structure of the paper.} Section~\ref{sec:prelims} briefly recalls the syntax and semantics of Deterministic Suffix-reading Automata (DSA). In Section~\ref{sec:multiport}, we extend DSA with the parallel operator, and in section~\ref{sec:multiport-outputs}, we add the feature to produce outputs. Finally in Section~\ref{sec:edt}, we describe the EDT notation and the translation from EDT to multi-port DSA with outputs. 